\pdfoutput=1
\documentclass[11pt]{article}

\usepackage{ACL2023}
\usepackage{times}
\usepackage{latexsym}
\usepackage[T1]{fontenc}

\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{fancyhdr}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{array}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE,RO]{EPFL -- Spring Semester 2026}
\fancyhead[RE,LO]{Modern NLP -- Open Project M2 Progress Report}
\rfoot{Page \thepage \hspace{1pt}}
\renewcommand{\headrulewidth}{1pt}

\setlength{\textfloatsep}{4pt}
\setlength{\intextsep}{4pt}
\setlength{\abovecaptionskip}{2pt}
\setlength{\belowcaptionskip}{0pt}

% Tighten section spacing for the 1-page brief.
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{6pt}{2pt}

%%%%%%%%%%%%%%%%%%% TITLE AND AUTHORS %%%%%%%%%%%%%%%%%%%

\title{M2 Progress Report -- Faithful RAG: Evaluating Citation Accuracy and Retrieval Robustness in Domain-Specific Scientific Literature}

\author{\normalfont
Elie Bruno | 355932 | \texttt{elie.bruno@epfl.ch} \\
Andrea Trugenberger | SCIPER | \texttt{andrea.trugenberger@epfl.ch} \\
Faruk Zahiragi\'c | SCIPER | \texttt{faruk.zahiragic@epfl.ch} \\
Yusif | SCIPER | \texttt{yusif@epfl.ch} \\
Team Faithful RAG \\
}

%%%%%%%%%%%%%%%%%%% PROGRESS REPORT %%%%%%%%%%%%%%%%%%%

\begin{document}
\maketitle
\thispagestyle{fancy}

\section{Progress Summary}

Three of the four proposal components run end-to-end on a 37-question gold benchmark of intellectual-property and innovation-policy papers. The retrieval ablation grew from four to sixteen configurations (new corpus-level SOTA); the faithfulness verifier exposes a 55--61\,pp gap between NLI and LLM-judge labels; the RAGAS pipeline pits chunked RAG against a 1M-token long-context baseline; the Corrective RAG hook is wired and awaits threshold ablations in M3.

\section{Gold Benchmark}
\label{sec:gold}

Each team member contributed $\sim$10 question-passage pairs across policy impact, methodology, comparison, factual extraction, and multi-hop reasoning. A pair encodes an atomic claim, the source \texttt{paper\_id}, character-span offsets on a line-feed-normalised \texttt{document.md}, and a verbatim quote. A validator byte-checks every span against its quote, and CI rejects fabricated concatenations. Inter-annotator agreement uses five pairs per annotator labelled \emph{supports}/\emph{contradicts}/\emph{unrelated}; six adversarial controls (q900--q905) populate the rejection cells and break the single-class Cohen-$\kappa$ paradox \citep{cohen1960kappa} that the first IAA round triggered.

\section{Retrieval Ablation}
\label{sec:retrieval}

The ablation crosses four embedder families (OpenAI \texttt{text-embedding-3-small}; BGE-M3 \citep{chen2024bgem3}; E5-large \citep{wang2024e5}; ColBERTv2 \citep{santhanam2022colbertv2}), two chunk granularities (coarse $\sim$2000\,chars, fine $\sim$300\,chars), and an optional ZeroEntropy cross-encoder reranker. Sixteen configurations share one \texttt{HybridRetriever} code path and report hit@$k$, MRR, and nDCG over $n{=}37$ paper-level gold pairs.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{lccc}
\toprule
config (\texttt{coarse\_rerank}) & hit@5 & hit@10 & MRR \\
\midrule
OpenAI \texttt{text-emb-3-small}     & 0.946          & 0.946          & 0.842 \\
BGE-M3                                & 0.946          & 0.946          & 0.842 \\
\textbf{E5-large}                     & \textbf{0.973} & \textbf{0.973} & \textbf{0.878} \\
ColBERTv2                             & 0.892          & 0.919          & 0.786 \\
\bottomrule
\end{tabular}
\caption{Best configuration per embedder family. Full 16-config grid lives in the repo \texttt{retrieval\_eval/} results.}
\label{tab:retrieval}
\end{table}

E5-large with the reranker wins every paper-level metric and beats OpenAI even without it ($+10.8$\,pp hit@10, $+8.6$\,pp MRR): a 335M-parameter open-source encoder edges out the closed-source baseline. ColBERTv2 trails E5-large by $17$\,pp MRR despite a more expressive per-token similarity; its MS-MARCO pretraining fails to transfer to scholarly text. ColBERTv2 is the only family that prefers fine chunks. Multi-hop questions saturate hit@10 at $1.000$ on coarse and fall to $0.667$--$0.833$ on fine across every family, locating the chunker as the dominant bottleneck.

\section{Faithfulness Verification}
\label{sec:faith}

The verifier extends the proposal's NLI pipeline (DeBERTa-v3-large-mnli, \citealp{he2021debertav3}) with an LLM-judge head. Three generators (GPT-4o-mini, Claude-3.5-haiku, DeepSeek-chat) answer the 24 answerable gold questions closed-book; both verifiers label each (claim, gold-passage) pair.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{lccr}
\toprule
generator & NLI & LLM-judge & gap \\
\midrule
gpt-4o-mini             & 0.750 & 0.138 & $+0.612$ \\
claude-3.5-haiku        & 0.750 & 0.173 & $+0.577$ \\
deepseek-chat           & 0.750 & 0.198 & $+0.552$ \\
\bottomrule
\end{tabular}
\caption{Fraction labelled \emph{faithful} by each verifier ($n{=}24$).}
\label{tab:nli-vs-judge}
\end{table}

NLI labels three-quarters of hallucinated closed-book answers as faithful at every provider; the LLM judge flags them. The 55--61\,pp gap is stable across generators and replicates a known NLI failure mode \citep{min-etal-2023-factscore,gao2023enabling}. NLI alone fails as a faithfulness gate on this benchmark.

\section{End-to-End RAG vs Long-Context}
\label{sec:ragas}

Two pipelines run through the four RAGAS metrics \citep{es2024ragas}: chunked RAG (the \S\ref{sec:retrieval} retriever, GPT-4o-mini) and a long-context baseline that loads every cited \texttt{document.md} into a 1M-token Gemini~2.5~Flash window.

\begin{table}[h]
\centering
\footnotesize
\begin{tabular}{lccr}
\toprule
metric & RAG & long-context & $\Delta$ \\
\midrule
faithfulness        & 0.806 & 0.975 & $+0.169$ \\
answer\_relevancy   & 0.662 & 0.799 & $+0.138$ \\
context\_precision  & 0.979 & 1.000 & $+0.021$ \\
context\_recall     & 0.792 & 1.000 & $+0.208$ \\
\bottomrule
\end{tabular}
\caption{RAGAS metrics, $n{=}8$ stratified subset.}
\label{tab:ragas}
\end{table}

Long-context wins every metric. The widest gaps fall on context\_recall ($+20.8$\,pp) and faithfulness ($+16.9$\,pp): chunked RAG hides $\sim$17\,pp of available factual support behind chunk boundaries. Two metrics point at one root cause and reinforce the chunker bottleneck from Table~\ref{tab:retrieval}.

\section{Limitations and M3 Plan}
\label{sec:m3}

Three scope ceilings remain: RAGAS at $n{=}8$, partial chunk-level coverage (16--20 of 37 queries), and a closed-book versus retrieved-context split between the two faithfulness verifiers. M3 delivers four fixes. (i) A no-gap sliding-window chunker: the multi-hop fine-chunk failures, partial chunk coverage, and the $+17$\,pp long-context advantage share one root cause. (ii) Domain-adaptation of ColBERTv2 on synthetic contrastive pairs from the gold set, sized to close the $17$\,pp MRR gap. (iii) The full 37-question RAGAS sweep across all 16 retriever configurations. (iv) Activation of the wired CRAG \citep{yan2024corrective} scorer as a 17th configuration aimed at the multi-hop \texttt{fine} failures isolated in \S\ref{sec:retrieval}.

\newpage
\bibliography{custom}
\bibliographystyle{acl_natbib}

\end{document}
